An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-05-24 EconGrowth.Rnw
Added nonbreaking spaces between Equation, Figure, Table, Section and the \ref{} reference.

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, with e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year. $A$ is a unitless technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$).\footnote{Dimensionless, indexed quantities are represented by lower-case symbols ($y$, $k$, $l$, $e$, $q$, $x$, and $u$), and dimensional quantities are represented by upper-case symbols ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$). One exception is the technolgical progress term $A$, which is dimensionless but written in upper-case by convention.} Neoclassical economic growth theory as expressed in the Cobb-Douglas production function makes no attempt to describe the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation \ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

As shown in Equation \ref{eq:CD} above, neoclassical economic theory does not include an explicit term for the role of energy in economic growth. However, \citet{Bashmakov:2007ek}, \citet{Hamilton:2009ui}, \citet{Murphy:2011jh}, \citet{Benes:2012tm}, \citet{Kumhof:2012ua}, and others indicate that the importance of energy for economic growth may far outweigh its cost share in the economy. That neoclassical economic growth theory ignores energy is, therefore, cause for concern, and inquiry into the proper approach for including energy as a factor of production in economic growth modeling is both relevant and important.

@@

The remainder of this paper is organized as follows. Section \ref{sec:Coordinates} discusses the coordinates of analysis described above. Section \ref{sec:Sources_of_Data} details sources of data and calculations to generate time series for economic output and factors of production. The procedures for and results from parameter estimation for the various production functions are discussed in Section \ref{sec:Parameter_Estimation}. Section \ref{sec:Implications} discusses implications of the parameter estimation results, and Section \ref{sec:Conclusion} concludes. An appendix provides details of all statistical results.

@@

Similar to Equation \ref{eq:CD}, the energy-augmented Cobb-Douglas production function is

@@

\citet{Ayres:2008cl} uses Equation \ref{eq:CDe} with exergy ($x$) and useful work ($u$) quantifications for $e$ to describe historic Japanese and American economies from 1900-2000. \citet{Froling:2009vo} uses an energy-augmented Cobb-Douglas production function with thermal energy quantification ($q$) to describe output for the global economy from 1800-1970. Fr\"{o}ling's model endogenizes labor with respect to fertility and mortality rates and separates thermal energy into two types: biomass energy and coal energy. \citet{Serrenho:2010vq} use the log transform of the energy-augmented Cobb-Douglas function with thermal energy quantification ($q$) to describe output for the EU-15 countries. Serrenho divides thermal energy into renewables, non-renewables, wastes, and imports.

@@

\noindent Equation \ref{eq:CES} is a CES production function with capital stock ($k$) and labor ($l$) factors of production. Equation \ref{eq:CESe} augments Equation \ref{eq:CES} with energy using a ($kl$)($e$) nesting structure, as is typical in the literature. Equation \ref{eq:CES} is a degenerate form of Equation \ref{eq:CESe} where $\delta = 1$. The elasticity of substitution between capital ($k$) and labor ($l$) is given by $\sigma_1 = \frac{1}{1+\rho_1}$, the elasticity of substitution between ($kl$) and ($e$) is given by $\sigma = \frac{1}{1+\rho}$, $\delta \in [0, 1]$ describes the importance of ($kl$) relative to ($e$), and $\gamma$ is a fitting parameter, expected to be close to unity.

@@

In contrast to both the Cobb-Douglas model (Equations \ref{eq:CD} and \ref{eq:CDe}) and the CES model (Equations \ref{eq:CES} and \ref{eq:CESe}), the LINEX model (Equation \ref{eq:LINEX}) accounts for efficiencies among capital, labor, and energy directly and, therefore, does not employ a time-dependent technological progress term ($A$). \citet{Kummel:2002tx} use the LINEX production function to describe economic output for the economies of the U.S., Japan, and Germany. \citet{Warr:2010ga} compare the LINEX production function with the Cobb-Douglas production function to assess economic growth for Austria, the United Kingdom, the United States, and Japan.

@@

We know of no statistically-rigorous comparisons of the predictive ability of each energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($Q$, $X$, and $U$) for developed economies and thermal energy ($Q$) and exergy ($X$) for all other economies.

@@

But, the nearly-complete focus on developed economies closes off the possibility of discovering any role that energy plays in the development process. In this paper we analyze at least two economies in each of the following four categories: Developed (technologically advanced with high standards of living), BRICS\footnote{Brazil, Russia, India, China, and South Africa} (rapid economic growth with standards of living below that of developed economies), Petrol (petroleum is a substantial source of GDP **** Martin: Is this the right way to describe the petrol economies? ****); and LDCs\footnote{Lesser Developed Countries} (living standards near subsistence levels **** Martin: Is this the right way to describe the LDC economies? ****). (See Table \ref{tab:Economy_Examples}.)

@@

To assess energy-augmented models of economic growth along the coordinates of analysis outlined in Section \ref{sec:Coordinates}, historical data for economic output ($Y$), capital stock ($K$), labor ($L$), thermal energy ($Q$), exergy ($X$), and useful work ($U$) are required. Subsections below describe sources for and calculations to obtain historical data. A final subsection presents the historical data in graphical form.

@@

To account for all energy input to an economy, we summed consumption data for many energy sources, including coal, oil and petroleum products, dry natural gas and natural gas plant liquids, nuclear power generation, and renewable energy sources (solar, wind, geothermal, biomass/fuel wood, and hydro). In the style of work by Warr and Ayres \citep{Ayres:2003ec, Ayres:2005kn, Warr:2010ht, Warr:2010ga, Warr:2012cg}, energy from human and animal muscle power was also included. Similar to the economic data discussed in Sections \ref{sec:Economic_Output_Data}--\ref{sec:Labor_Data}, we generated energy data for developed countries from 1980--2011, with the exception of useful work data (1980--2000). For all other countries, we generated thermal energy and exergy data for 1991--2011. Details on each energy quantifications are given in subsections below.

@@

For renewable and nuclear energy sources, yearly electricity generation (in BTU/year) was available directly from the \citet{EIA:2013a} for all economies, so Equation \ref{eq:thermal_energy} was unnecessary.

@@

Exergy ($X$), expressed in TJ/year, was calculated by multiplying the thermal energy ($Q$) values discussed in Section \ref{sec:Q_Data} by a unitless exergy-to-thermal-energy allocation factor ($f_{exergy-allocation}$) shown in Table \ref{tab:Exergy_Allocation}.

@@

\noindent $\eta_{food \rightarrow work, animals}$ and $\eta_{food \rightarrow work, workers}$ are animal and worker efficiency of converting food into energy, respectively. The animal feed-to-work conversion efficiency ($\eta_{feed \rightarrow work, animals}$ = 0.4) was taken from \citet{Warr:2012a} and assumed equal across all economies. The food-to-work efficiency for workers ($\eta_{food \rightarrow work, workers}$ = 0.2) was taken from \citet[Table~B.1]{Ayres:2010ug}. Following \citet[p.~315]{Ayres:2010ug} and methodology from \citet{Warr:2012a}, an additional factor ($f_{intake/wholesale}$) was included to account for ``wastage and losses of food in the household, e.g. during storage, in preparation and cooking, as plate-waste, or quantities fed to domestic animals and pets'' \citep{FAO:2013b}. Values of $f_{intake/wholesale}$ were calculated from \citet[Table~3.3 for wholesale and Table~3.4 for intake]{Wirsenius:2000vi}. $f_{intake/wholesale}$ varies by country (Table \ref{tab:f_intake_wholesale}). A similar intake-to-wholesale ratio could also be applied to animal work, but data to calculate the ratio was unavailable from \citet{Wirsenius:2000vi}.\footnote{The error in exergy caused by neglecting $f_{intake/wholesale}$ for animals is very small, because animal muscle exergy is a small fraction of the total energy supply to any economy, with a maximum of 0.41\% for China in 1992. It could be argued that $f_{intake/wholesale}$ for animals is likely to be closer to 1.0 than $f_{intake/wholesale}$ for humans.}

@@

\noindent where the subscript $i$ indicates the energy enduse, $\varepsilon_i$ is the efficiency of converting exergy into useful energy for each enduse, and $\phi_i$ gives the fraction of exergy input to an economy that goes toward the i$^\mathrm{th}$ enduse.

Values of $\varepsilon_i$ and $\phi_i$ were calculated from the REXS database supplied by \citet{Warr:2012a}. Values of $X$ were calculated by methods discussed in Section \ref{sec:X_Data}. For most countries, values of $\phi_i$ are not constant with time, as economic sectors with different distributions of raw exergy inputs grow at different rates. \citeapos{Warr:2012a} REXS database contains information for developed economies (United States, United Kingdom, and Japan) from 1900--2000, and we used data from the 1980--2000 timeframe. Extending the REXS database to the BRICS, petrol, and LDC economies is beyond the scope of this work, because longitudinal data for the efficiency of different end uses ($\varepsilon_i$) and the distribution of exergy sources to end uses ($\phi_i$) are extremely difficult to find.

@@

Figure \ref{fig:Historical_Lattice_Graph} presents historical data, including indexed values for output ($y$), capital stock ($k$), labor ($l$), and the various energy quantifications ($q$, $x$, and $u$) for 1980--2011 (developed economies) and 1991--2011 (other economies). Factors of production track each other closely, and high covariances among these variables can be seen in Tables \ref{tab:Covariance_US} through \ref{tab:Covariance_ZM}. **** Discuss the covariances in more detail after we finalize the covariances for U. ****

@@

Figure \ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution in the economy (such as China's rapid increase of coal and natural gas consumption relative to petroleum since 2000). The growth rate of $u$ is significantly different from $q$ and $x$ for most countries, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations \ref{eq:CD} -- \ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.

@@

Each of the economic growth models discussed in Section \ref{sec:Prod_Func_Forms} has parameters which must be estimated for each economy (see Section \ref{sec:Economy_Types}) and each energy quantification (see Section \ref{sec:Quantification_of_Energy}) via least-squares methods. The following subsections discuss parameter estimation methods interpret the results.

@@

The Cobb-Douglas model without energy is given by Equation \ref{eq:CD}. The Cobb-Douglas production function was reparameterized as

@@

\noindent to ensure $\alpha + \beta = 1$ for constant returns to scale. By constraining $\alpha \in [0,1]$, we ensure that $\beta \in [0,1]$ also. **** Is this truly how we implemented the constraints? Finish after we actually implement the constraints. **** $\lambda$ and $\alpha$ were estimated by the \citet{R} function \texttt{nls}.  The value of $\beta$ can be found with $\beta = 1 - \alpha$. Figure \ref{fig:CD_Params_Graph} shows values of and 95\% confidence intervals for $\lambda$ and $\alpha$. Appendix \ref{sec:Model_Parameters} contains tables with numerical values of model parameters and confidence intervals.

@@

To qualitatively assess whether augmenting the Cobb-Douglas model with energy will be beneficial, we created partial residual plots as shown in Figures \ref{fig:CDPartialResidualsQ}--\ref{fig:CDPartialResidualsU} below. If a monotonic trend is observed on the graphs, we expect that energy augmentation will be beneficial. However, no strong monotonic tends are observed for any country, with the exception, perhaps, of the United Kingdom, whose residuals are, on average, above 0.

@@

To estimate the parameters $\lambda$, $\alpha$, $\beta$, and $\gamma$ in the energy-augmented Cobb-Douglas model, Equation \ref{eq:CDe} was reparameterized as

@@

The Cobb-Douglas with energy parameter estimation process is challenging. To demonstrate the characteristics of the parameter estimation process, we performed a grid search over values of $\gamma$, wherein we re-estimated values of $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$. Figure \ref{fig:CDeGridLattice} shows the sum of squared errors ($SSE$), $\alpha$, and $\beta$ as a function of the fixed value of $\gamma$ with thermal energy ($q$). For most economies, the minimum $SSE$ occurs at a boundary of $\alpha$, $\beta$, or $\gamma$. The parameter estimation algorithm has difficulty assessing confidence intervals for $\alpha$, $\beta$, and $\gamma$, because the constrained model behaves very differently on either side of the constraint.

@@

Figure \ref{fig:CDq_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $q$).

@@

The Cobb-Douglas (with exergy) parameters are given in Table \ref{tab:CD_Parameters_With_X}.

Figure \ref{fig:CDx_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $x$).

@@

Figure \ref{fig:CDu_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $u$).

@@

Figure \ref{fig:CD_GDP_Lattice_Graph} compares predictions from the Cobb-Douglas models (without energy, with $e=q$, with $e=x$, and with $e=u$) to historical data.

@@

The CES model without and with energy is given by Equations \ref{eq:CES} and \ref{eq:CESe}, respectively. The \texttt{R} \citep{R} package \texttt{micEconCES} \citep{Henningsen:2011td} was used to estimate parameters $\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, and $\rho$. **** More details here about which algorithms were used, etc. Complete after our work this summer. **** Elasticities of substitution $\sigma_1$ and $\sigma$ (see Section \ref{sec:CES}) were estimated by the delta method (**** is this correct? Ask Pruim. ****).

@@

Figure \ref{fig:CES_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (without energy).

@@

Similar to the analysis of the Cobb-Douglas model in Section \ref{sec:CD_Parameter_Estimation}, we make partial residual plots for the CES model (without energy) in Figures \ref{fig:CESPartialResidualsQ} - \ref{fig:CESPartialResidualsU} below. Again, no strong monotonic trend is observed, suggesting the energy-augmentation of the CES model will not significantly improve the model.

@@

Figure \ref{fig:CESq_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $q$).

@@

Figure \ref{fig:CESx_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $x$).

@@

Figure \ref{fig:CESu_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $u$).

@@

Figure \ref{fig:CES_GDP_Lattice_Graph} compares CES model predictions to historical data.

@@

The key energy parameters for the Cobb-Douglas and CES models are $\gamma$, $\delta$, and $\sigma$. The parameter $\gamma$ in the Cobb-Douglas model (Equation \ref{eq:CDe}) indicates the importance of energy for economic growth. When $\gamma \rightarrow 0$, energy is not a significant contributor to economic growth for an economy. Conversely, when $\gamma \rightarrow 1$, energy is the primary contributor to economic growth. However, if the 95\% confidence interval for $\gamma$ for any given economy extends to 0.0, we cannot say with high confidence whether or not energy makes a meaningful contribution to economic growth for that economy. On a graph of the width of the 95\% confidence interval for $\gamma$ ($CI_{95\%,\gamma}$) vs. $\gamma$, where each economy is represented by a point (its two-letter abbreviation; see Table \ref{tab:Economy_Examples}), the locus of points where the 95\% confidence interval for $\gamma$ crosses 0 is a diagonal line. If the combination of $\gamma$ and its confidence interval for a given economy is below and to the right of the line, we have high confidence that we can say whether or not energy is a significant contributor to economic growth. Thus, only when $\gamma$ is below the line and small (i.e., close to 0) for an economy, can we say with confidence that energy is an insignificant contributor to economic growth. Only when $\gamma$ is below the line and large (i.e., close to 1) for an economy can we say that energy is a significant contributor to economic growth, according to the Cobb-Douglas model.

The left column of graphs in Figure \ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\gamma}$ vs. $\gamma$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that most economies are above and to the left of the line, while only a few are below and to the right.

>>>>

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, with e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year. $A$ is a unitless technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$).\footnote{Dimensionless, indexed quantities are represented by lower-case symbols ($y$, $k$, $l$, $e$, $q$, $x$, and $u$), and dimensional quantities are represented by upper-case symbols ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$). One exception is the technolgical progress term $A$, which is dimensionless but written in upper-case by convention.} Neoclassical economic growth theory as expressed in the Cobb-Douglas production function makes no attempt to describe the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation~\ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

As shown in Equation~\ref{eq:CD} above, neoclassical economic theory does not include an explicit term for the role of energy in economic growth. However, \citet{Bashmakov:2007ek}, \citet{Hamilton:2009ui}, \citet{Murphy:2011jh}, \citet{Benes:2012tm}, \citet{Kumhof:2012ua}, and others indicate that the importance of energy for economic growth may far outweigh its cost share in the economy. That neoclassical economic growth theory ignores energy is, therefore, cause for concern, and inquiry into the proper approach for including energy as a factor of production in economic growth modeling is both relevant and important.

@@

The remainder of this paper is organized as follows. Section~\ref{sec:Coordinates} discusses the coordinates of analysis described above. Section~\ref{sec:Sources_of_Data} details sources of data and calculations to generate time series for economic output and factors of production. The procedures for and results from parameter estimation for the various production functions are discussed in Section~\ref{sec:Parameter_Estimation}. Section~\ref{sec:Implications} discusses implications of the parameter estimation results, and Section~\ref{sec:Conclusion} concludes. An appendix provides details of all statistical results.

@@

Similar to Equation~\ref{eq:CD}, the energy-augmented Cobb-Douglas production function is

@@

\citet{Ayres:2008cl} uses Equation~\ref{eq:CDe} with exergy ($x$) and useful work ($u$) quantifications for $e$ to describe historic Japanese and American economies from 1900-2000. \citet{Froling:2009vo} uses an energy-augmented Cobb-Douglas production function with thermal energy quantification ($q$) to describe output for the global economy from 1800-1970. Fr\"{o}ling's model endogenizes labor with respect to fertility and mortality rates and separates thermal energy into two types: biomass energy and coal energy. \citet{Serrenho:2010vq} use the log transform of the energy-augmented Cobb-Douglas function with thermal energy quantification ($q$) to describe output for the EU-15 countries. Serrenho divides thermal energy into renewables, non-renewables, wastes, and imports.

@@

\noindent Equation~\ref{eq:CES} is a CES production function with capital stock ($k$) and labor ($l$) factors of production. Equation~\ref{eq:CESe} augments Equation~\ref{eq:CES} with energy using a ($kl$)($e$) nesting structure, as is typical in the literature. Equation~\ref{eq:CES} is a degenerate form of Equation~\ref{eq:CESe} where $\delta = 1$. The elasticity of substitution between capital ($k$) and labor ($l$) is given by $\sigma_1 = \frac{1}{1+\rho_1}$, the elasticity of substitution between ($kl$) and ($e$) is given by $\sigma = \frac{1}{1+\rho}$, $\delta \in [0, 1]$ describes the importance of ($kl$) relative to ($e$), and $\gamma$ is a fitting parameter, expected to be close to unity.

@@

In contrast to both the Cobb-Douglas model (Equations~\ref{eq:CD} and \ref{eq:CDe}) and the CES model (Equations~\ref{eq:CES} and \ref{eq:CESe}), the LINEX model (Equation~\ref{eq:LINEX}) accounts for efficiencies among capital, labor, and energy directly and, therefore, does not employ a time-dependent technological progress term ($A$). \citet{Kummel:2002tx} use the LINEX production function to describe economic output for the economies of the U.S., Japan, and Germany. \citet{Warr:2010ga} compare the LINEX production function with the Cobb-Douglas production function to assess economic growth for Austria, the United Kingdom, the United States, and Japan.

@@

We know of no statistically-rigorous comparisons of the predictive ability of each energy quantification in the various types of production functions discussed in Section~\ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($Q$, $X$, and $U$) for developed economies and thermal energy ($Q$) and exergy ($X$) for all other economies.

@@

But, the nearly-complete focus on developed economies closes off the possibility of discovering any role that energy plays in the development process. In this paper we analyze at least two economies in each of the following four categories: Developed (technologically advanced with high standards of living), BRICS\footnote{Brazil, Russia, India, China, and South Africa} (rapid economic growth with standards of living below that of developed economies), Petrol (petroleum is a substantial source of GDP **** Martin: Is this the right way to describe the petrol economies? ****); and LDCs\footnote{Lesser Developed Countries} (living standards near subsistence levels **** Martin: Is this the right way to describe the LDC economies? ****). (See Table~\ref{tab:Economy_Examples}.)

@@

To assess energy-augmented models of economic growth along the coordinates of analysis outlined in Section~\ref{sec:Coordinates}, historical data for economic output ($Y$), capital stock ($K$), labor ($L$), thermal energy ($Q$), exergy ($X$), and useful work ($U$) are required. Subsections below describe sources for and calculations to obtain historical data. A final subsection presents the historical data in graphical form.

@@

To account for all energy input to an economy, we summed consumption data for many energy sources, including coal, oil and petroleum products, dry natural gas and natural gas plant liquids, nuclear power generation, and renewable energy sources (solar, wind, geothermal, biomass/fuel wood, and hydro). In the style of work by Warr and Ayres \citep{Ayres:2003ec, Ayres:2005kn, Warr:2010ht, Warr:2010ga, Warr:2012cg}, energy from human and animal muscle power was also included. Similar to the economic data discussed in Sections~\ref{sec:Economic_Output_Data}--\ref{sec:Labor_Data}, we generated energy data for developed countries from 1980--2011, with the exception of useful work data (1980--2000). For all other countries, we generated thermal energy and exergy data for 1991--2011. Details on each energy quantifications are given in subsections below.

@@

For renewable and nuclear energy sources, yearly electricity generation (in BTU/year) was available directly from the \citet{EIA:2013a} for all economies, so Equation~\ref{eq:thermal_energy} was unnecessary.

@@

Exergy ($X$), expressed in TJ/year, was calculated by multiplying the thermal energy ($Q$) values discussed in Section~\ref{sec:Q_Data} by a unitless exergy-to-thermal-energy allocation factor ($f_{exergy-allocation}$) shown in Table~\ref{tab:Exergy_Allocation}.

@@

\noindent $\eta_{food \rightarrow work, animals}$ and $\eta_{food \rightarrow work, workers}$ are animal and worker efficiency of converting food into energy, respectively. The animal feed-to-work conversion efficiency ($\eta_{feed \rightarrow work, animals}$ = 0.4) was taken from \citet{Warr:2012a} and assumed equal across all economies. The food-to-work efficiency for workers ($\eta_{food \rightarrow work, workers}$ = 0.2) was taken from \citet[Table~B.1]{Ayres:2010ug}. Following \citet[p.~315]{Ayres:2010ug} and methodology from \citet{Warr:2012a}, an additional factor ($f_{intake/wholesale}$) was included to account for ``wastage and losses of food in the household, e.g. during storage, in preparation and cooking, as plate-waste, or quantities fed to domestic animals and pets'' \citep[\texttt{notes} tab]{FAO:2013b}. Values of $f_{intake/wholesale}$ were calculated from \citet[Table~3.3 for wholesale and Table~3.4 for intake]{Wirsenius:2000vi}. $f_{intake/wholesale}$ varies by country as shown in Table~\ref{tab:f_intake_wholesale}. A similar intake-to-wholesale ratio could also be applied to animal work, but data to calculate the ratio was unavailable from \citet{Wirsenius:2000vi}, and $f_{intake/wholesale,animals}$ is assumed to be unity.\footnote{The error in total exergy input to any economy caused by assuming $f_{intake/wholesale,animals} = 1.0$ is very small, because animal muscle exergy is a small fraction of the total exergy supply to any economy, with a maximum of 0.41\% for China in 1992. It could also be argued that $f_{intake/wholesale, animals}$ is likely to be closer to unity than $f_{intake/wholesale, workers}$.}

@@

\noindent where the subscript $i$ indicates the energy enduse, $\varepsilon_i$ is the efficiency of converting exergy into useful energy for each enduse, and $\phi_i$ gives the fraction of exergy input to an economy that goes toward the $i^\mathrm{th}$ enduse.

Values of $\varepsilon_i$ and $\phi_i$ were calculated from the REXS database supplied by \citet{Warr:2012a}. Values of $X$ were calculated by methods discussed in Section~\ref{sec:X_Data}. For most countries, values of $\phi_i$ are not constant with time, as economic sectors with different distributions of raw exergy inputs grow at different rates. \citeapos{Warr:2012a} REXS database contains information for developed economies (United States, United Kingdom, and Japan) from 1900--2000, and we used data from the 1980--2000 timeframe. Extending the REXS database to the BRICS, petrol, and LDC economies is beyond the scope of this work, because longitudinal data for the efficiency of different end uses ($\varepsilon_i$) and the distribution of exergy sources to end uses ($\phi_i$) are extremely difficult to find.

@@

Figure~\ref{fig:Historical_Lattice_Graph} presents historical data, including indexed values for output ($y$), capital stock ($k$), labor ($l$), and the various energy quantifications ($q$, $x$, and $u$) for 1980--2011 (developed economies) and 1991--2011 (other economies). Factors of production track each other closely, and high covariances among these variables can be seen in Tables~\ref{tab:Covariance_US}--\ref{tab:Covariance_ZM}. **** Discuss the covariances in more detail after we finalize the covariances for U. ****

@@

Figure~\ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution in the economy (such as China's rapid increase of coal and natural gas consumption relative to petroleum since 2000). The growth rate of $u$ is significantly different from $q$ and $x$ for most countries, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations~\ref{eq:CD}--\ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.

@@

Each of the economic growth models discussed in Section~\ref{sec:Prod_Func_Forms} has parameters which must be estimated for each economy (see Section~\ref{sec:Economy_Types}) and each energy quantification (see Section~\ref{sec:Quantification_of_Energy}) via least-squares methods. The following subsections discuss parameter estimation methods interpret the results.

@@

The Cobb-Douglas model without energy is given by Equation~\ref{eq:CD}. The Cobb-Douglas production function was reparameterized as

@@

\noindent to ensure $\alpha + \beta = 1$ for constant returns to scale. By constraining $\alpha \in [0,1]$, we ensure that $\beta \in [0,1]$ also. **** Is this truly how we implemented the constraints? Finish after we actually implement the constraints. **** $\lambda$ and $\alpha$ were estimated by the \citet{R} function \texttt{nls}.  The value of $\beta$ can be found with $\beta = 1 - \alpha$. Figure~\ref{fig:CD_Params_Graph} shows values of and 95\% confidence intervals for $\lambda$ and $\alpha$. Appendix~\ref{sec:Model_Parameters} contains tables with numerical values of model parameters and confidence intervals.

@@

To qualitatively assess whether augmenting the Cobb-Douglas model with energy will be beneficial, we created partial residual plots as shown in Figures~\ref{fig:CDPartialResidualsQ}--\ref{fig:CDPartialResidualsU} below. If a monotonic trend is observed on the graphs, we expect that energy augmentation will be beneficial. However, no strong monotonic tends are observed for any country, with the exception, perhaps, of the United Kingdom, whose residuals are, on average, above 0.

@@

To estimate the parameters $\lambda$, $\alpha$, $\beta$, and $\gamma$ in the energy-augmented Cobb-Douglas model, Equation~\ref{eq:CDe} was reparameterized as

@@

The Cobb-Douglas with energy parameter estimation process is challenging. To demonstrate the characteristics of the parameter estimation process, we performed a grid search over values of $\gamma$, wherein we re-estimated values of $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$. Figure~\ref{fig:CDeGridLattice} shows the sum of squared errors ($SSE$), $\alpha$, and $\beta$ as a function of the fixed value of $\gamma$ with thermal energy ($q$). For most economies, the minimum $SSE$ occurs at a boundary of $\alpha$, $\beta$, or $\gamma$. The parameter estimation algorithm has difficulty assessing confidence intervals for $\alpha$, $\beta$, and $\gamma$, because the constrained model behaves very differently on either side of the constraint.

@@

Figure~\ref{fig:CDq_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $q$).

@@

The Cobb-Douglas (with exergy) parameters are given in Table~\ref{tab:CD_Parameters_With_X}.

Figure~\ref{fig:CDx_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $x$).

@@

Figure~\ref{fig:CDu_Params_Graph} shows values and 95\% confidence intervals for the parameters for the Cobb-Douglas model (with $u$).

@@

Figure~\ref{fig:CD_GDP_Lattice_Graph} compares predictions from the Cobb-Douglas models (without energy, with $e=q$, with $e=x$, and with $e=u$) to historical data.

@@

The CES model without and with energy is given by Equations~\ref{eq:CES} and \ref{eq:CESe}, respectively. The \texttt{R} \citep{R} package \texttt{micEconCES} \citep{Henningsen:2011td} was used to estimate parameters $\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, and $\rho$. **** More details here about which algorithms were used, etc. Complete after our work this summer. **** Elasticities of substitution $\sigma_1$ and $\sigma$ (see Section~\ref{sec:CES}) were estimated by the delta method (**** is this correct? Ask Pruim. ****).

@@

Figure~\ref{fig:CES_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (without energy).

@@

Similar to the analysis of the Cobb-Douglas model in Section~\ref{sec:CD_Parameter_Estimation}, we make partial residual plots for the CES model (without energy) in Figures~\ref{fig:CESPartialResidualsQ}--\ref{fig:CESPartialResidualsU} below. Again, no strong monotonic trend is observed, suggesting the energy-augmentation of the CES model will not significantly improve the model.

@@

Figure~\ref{fig:CESq_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $q$).

@@

Figure~\ref{fig:CESx_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $x$).

@@

Figure~\ref{fig:CESu_Params_Graph} shows values and 95\% confidence intervals for the parameters for the CES model (with $u$).

@@

Figure~\ref{fig:CES_GDP_Lattice_Graph} compares CES model predictions to historical data.

@@

The key energy parameters for the Cobb-Douglas and CES models are $\gamma$, $\delta$, and $\sigma$. The parameter $\gamma$ in the Cobb-Douglas model (Equation~\ref{eq:CDe}) indicates the importance of energy for economic growth. When $\gamma \rightarrow 0$, energy is not a significant contributor to economic growth for an economy. Conversely, when $\gamma \rightarrow 1$, energy is the primary contributor to economic growth. However, if the 95\% confidence interval for $\gamma$ for any given economy extends to 0.0, we cannot say with high confidence whether or not energy makes a meaningful contribution to economic growth for that economy. On a graph of the width of the 95\% confidence interval for $\gamma$ ($CI_{95\%,\gamma}$) vs. $\gamma$, where each economy is represented by a point (its two-letter abbreviation; see Table~\ref{tab:Economy_Examples}), the locus of points where the 95\% confidence interval for $\gamma$ crosses 0 is a diagonal line. If the combination of $\gamma$ and its confidence interval for a given economy is below and to the right of the line, we have high confidence that we can say whether or not energy is a significant contributor to economic growth. Thus, only when $\gamma$ is below the line and small (i.e., close to 0) for an economy, can we say with confidence that energy is an insignificant contributor to economic growth. Only when $\gamma$ is below the line and large (i.e., close to 1) for an economy can we say that energy is a significant contributor to economic growth, according to the Cobb-Douglas model.

The left column of graphs in Figure~\ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\gamma}$ vs. $\gamma$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that most economies are above and to the left of the line, while only a few are below and to the right.

For the CES model (Equation~\ref{eq:CESe}), the parameter $\delta$ indicates the importance of energy for economic growth. When $\delta \rightarrow 1$, energy is an insignificant contributor to economic growth, and when $\delta \rightarrow 0$, energy is the primary contributor to economic growth for an economy. However, if the 95\% confidence interval for $\delta$ extends to 1, we cannot say with high confidence whether or not energy makes a meaningful contribution to economic growth for that economy. In a manner similar to $\gamma$ for the Cobb-Douglas model, we note that if the combination of $\delta$ and its confidence interval for a given economy is below and to the left of the line, we have high confidence that we can say whether or not energy is a significant contributor to economic growth. Thus, only when $\delta$ is below the line and large (i.e., close to 1) for an economy, can we say with confidence that energy is an insignificant contributor to economic growth. Only when $\delta$ is below the line and small (i.e., close to 0) for an economy can we say that energy is a significant contributor to economic growth, according to the CES model.

The middle column of graphs in Figure~\ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\delta}$ vs. $\delta$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that most economies are above and to the right of the line, while only a few are below and to the left.
